\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2017-09-18 paper/casebase_jss.tex
combine vignette with talk

\begin{itemize}
\tightlist
\item
  Competing risk example
\item
  Comparison with Cox and Fine-Gray
\end{itemize}

@@

>>>>

The next example shows how case-base sampling can also be used in the
context of a competing risk analysis. For illustrative purposes, we will
use the same data that was used in Scrucca \emph{et al}
\citeyearpar{scrucca2010regression}. The data was downloaded from the
main author's website, and it is also available as part of the
\pkg{casebase} package.

\begin{CodeChunk}

\begin{CodeInput}
R> library(casebase)
R> data(bmtcrr)
\end{CodeInput}
\end{CodeChunk}

The data contains information on 177 patients who received a stem-cell
transplant for acute leukemia. The event of interest is relapse, but
other competing causes (e.g.~transplant-related death) were also
recorded Several covariates were also captured at baseline: sex, disease
type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL
and AML, respectively), disease phase at transplant (Relapse, CR1, CR2,
CR3), source of stem cells (bone marrow and peripheral blood, coded as
BM+PB, or only peripheral blood, coded as PB), and age. A summary of
these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
We note that the statistical summaries were generated differently for
different variable types: for continuous variables, we gave the range,
followed by the mean and standard deviation; for categorical variables,
we gave the counts for each category.

\begin{table}[ht]
\centering
\begin{tabular}{lll}
  \hline
Variable & Description & Statistical summary \\ 
  \hline
Sex & Sex & M=Male (100) \\ 
   &  & F=Female (77) \\ 
  D & Disease & ALL (73) \\ 
   &  & AML (104) \\ 
  Phase & Phase & CR1 (47) \\ 
   &  & CR2 (45) \\ 
   &  & CR3 (12) \\ 
   &  & Relapse (73) \\ 
  Source & Type of transplant & BM+PB (21) \\ 
   &  & PB (156) \\ 
  Age & Age of patient (years) & 4–62 \\ 
   &  & 30.47 (13.04) \\ 
  Ftime & Failure time (months) & 0.13–131.77 \\ 
   &  & 20.28 (30.78) \\ 
  Status & Status indicator & 0=censored (46) \\ 
   &  & 1=relapse (56) \\ 
   &  & 2=competing event (75) \\ 
   \hline
\end{tabular}
\caption{Baseline characteristics of patients in the stem-cell transplant study.}
\label{tab:table1bmtcrr}
\end{table}

In order to try and visualize the incidence density of relapse, we can
look at the corresponding population-time plot. In Figure
\ref{fig:compPop1}, failure times associated with relapse are
highlighted on the plot using red points, while Figure
\ref{fig:compPop2} provides a similar population-time plot for competing
events.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{casebase_jss_files/figure-latex/compPop1-1}

}

\caption{\label{fig:compPop1}Population-time plot for the stem-cell transplant study. The points represent the event of interest (i.e., relapse).}\label{fig:compPop1}
\end{figure}
\end{CodeChunk}

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{casebase_jss_files/figure-latex/compPop2-1}

}

\caption{\label{fig:compPop2}Population-time plot for the stem-cell transplant study. The points represent the competing events.}\label{fig:compPop2}
\end{figure}
\end{CodeChunk}

Our main objective is to compute the absolute risk of relapse for a
given set of covariates. First, we fit a smooth hazard to the data; for
the sake of this example, we opted for a linear term for time:

\begin{CodeChunk}

\begin{CodeInput}
R> model_cb <- fitSmoothHazard(
R+     Status ~ ftime + Sex + D + Phase + Source + Age, 
R+     data = bmtcrr, 
R+     ratio = 100, 
R+     time = "ftime")
\end{CodeInput}
\end{CodeChunk}

From the fit object, we can extract both the hazard ratios and their
corresponding confidence intervals:

\begin{CodeChunk}

\begin{tabular}{l|r|l}
\hline
Covariates & HR & 95\% CI\\
\hline
Sex & 0.71 & (0.41, 1.24)\\
\hline
Disease & 0.52 & (0.28, 0.93)\\
\hline
Phase (CR2 vs. CR1) & 1.17 & (0.47, 2.93)\\
\hline
Phase (CR3 vs. CR1) & 1.78 & (0.46, 6.91)\\
\hline
Phase (Relapse vs. CR1) & 4.45 & (2.07, 9.57)\\
\hline
Source & 1.64 & (0.54, 5)\\
\hline
Age & 0.99 & (0.97, 1.02)\\
\hline
\end{tabular}

\end{CodeChunk}

As we can see, the only significant hazard ratio is the one associated
with the phase of the disease at transplant. More precisely, being in
relapse at transplant is associated with a hazard ratio of 3.92 when
compared to CR1.

Given our estimate of the hazard function, we can compute the absolute
risk curve for a fixed covariate profile. We performed this computation
for a 35 year old woman who received a stem-cell transplant from
peripheral blood at relapse. We compared the absolute risk curve for
such a woman with acute lymphoblastic leukemia with that for a similar
woman with acute myeloblastic leukemia. Figure \ref{fig:compAbsrisk}
shows these two curves as a function of time. This figure also shows the
Kaplan-Meier estimate fitted to the two disease groups (ignoring the
other covariates).

\begin{CodeChunk}

\begin{CodeInput}
R> # Pick 100 equidistant points between 0 and 60 months
R> time_points <- seq(0, 60, length.out = 100)
R> 
R> # Data.frame containing risk profile
R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
R+                                      levels = levels(bmtcrr[,"Sex"])),
R+                       "D" = c("ALL", "AML"),
R+                       "Phase" = factor(c("Relapse", "Relapse"), 
R+                                        levels = levels(bmtcrr[,"Phase"])),
R+                       "Age" = c(35, 35),
R+                       "Source" = factor(c("PB", "PB"), 
R+                                         levels = levels(bmtcrr[,"Source"])))
R> 
R> # Estimate absolute risk curve
R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
R+                         method = "montecarlo", newdata = newdata)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-6-1}

}

\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:unnamed-chunk-6}
\end{figure}
\end{CodeChunk}

@@

\section{Discussion}\label{discussion}

\bibliography{casebase_references.bib}
